Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-12-04 manu-draft.qmd
esco csl

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary"
bibliography: refs.bib
# csl: limnology-and-oceanography.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true

@@

An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale. Our choice to use 30$^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed, as shown by the mixed-effects linear models for each bay segment. That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships. These thresholds are likely not ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003]. Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values (i.e., higher temperature, lower salinity), conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses. This does not suggest that these factors are unimportant, both currently and in the future. Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity suggest that conditions are very likely to become more stressful in the futre. Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were often not captured by the daily predictions (see supplement). Regardless of the above limitations, the role of climate stressors on seagrass change is likely to continue in the future as suggested by our results.

Additional limitations of our model may relate to an incomplete description of factors influencing seagrass growth. This relates to the inclusion of additional drivers as well as issues related to correlation/causation. For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth \[[@greening2014a; @sherwood2017a]. As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity. Our use of time period (pre-/post-2016) as a categorical variable partially addressed this issue. Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass. Thus, time period was necessary to control for these confounding relationships. Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay (OTB in particular where the most losses have been seen) \[@tbep0696; @Beck20a\]. This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity. A final consideration for our models relates to how seagrasses may influence their environment, particularly for the EPC and PDEM datasets where temperature and salinity were measured at the same locations as seagrass. For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses. This explanation cannot be ruled out with the existing datasets, although the trend analyses showing the consistent long-term changes and our models suggest that climate-related stressors driving seagrass loss is a more likely scenario.

@@

Additional biotic factors could be influencing regional patterns in seagrass growth. In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern [@jei2022; @scolaro2023]. Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay \[@Beck20b\]. Competitive differences between seagrasses and macroalgae are poorly understood in these systems \[but see @bell1997\], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae. Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth \[@beck2022; @scolaro2023; @tomasko2023\], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years. Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018]. For example, the parasitic slime mold *Labryinthula* spp. that causes seagrass wasting disease has been known to infect *Thalassia testudinium* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects. Existing research has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].

>>>>

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary"
bibliography: refs.bib
csl: estuaries-and-coasts.csl
author:
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true

@@

An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale. Our choice to use 30$^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed, as shown by the mixed-effects linear models for each bay segment. That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships. These thresholds are likely not ecologically significant since most of the reported values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003]. Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values (i.e., higher temperature, lower salinity), conditions in Tampa Bay in recent years are generally within the ecological niche for seagrasses. This does not suggest that these factors are unimportant, both currently and in the future. Our models suggested that temperature and salinity are potentially associated with seagrass loss and the consistent long-term trend in temperature and salinity suggest that conditions are very likely to become more stressful in the future. Moreover, the GAMs used to model continuous temperature and salinity at each EPC monitoring station likely did not fully describe the current extent of stress given that extreme low or high values were often not captured by the daily predictions (see supplement). Regardless of the above limitations, the role of climate stressors on seagrass change is likely to continue in the future as suggested by our results.

Additional limitations of our model may relate to an incomplete description of factors influencing seagrass growth. This relates to the inclusion of additional drivers as well as issues related to correlation/causation. For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth [@greening2014a; @sherwood2017a]. As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity. Our use of time period (pre-/post-2016) as a categorical variable partially addressed this issue. Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass. Thus, time period was necessary to control for these confounding relationships. Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay (OTB in particular where the most losses have been seen) [@tbep0696; @Beck20a]. This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity. A final consideration for our models relates to how seagrasses may influence their environment, particularly for the EPC and PDEM datasets where temperature and salinity were measured at the same locations as seagrass. For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses. This explanation cannot be ruled out with the existing datasets, although the trend analyses showing the consistent long-term changes and our models suggest that climate-related stressors driving seagrass loss is a more likely scenario.

@@

Additional biotic factors could be influencing regional patterns in seagrass growth. In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern [@jei2022; @scolaro2023]. Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay [@Beck20b]. Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae. Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth [@beck2022; @scolaro2023; @tomasko2023], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years. Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018]. For example, the parasitic slime mold *Labryinthula* spp. that causes seagrass wasting disease has been known to infect *Thalassia testudinium* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects. Existing research has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].
